### 2020-07-02 chapters/background.tex
Restivo feedback applied

@@

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch datasets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes). Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop \cite{borthakur2007hadoop} \cite{Hadoop} \cite{ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark} \cite{Spark} was released as the industry accepted successor.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

Stream procesisng, on the other hand, refers to the processing of data streams. A data stream is essentially a continuous or unbounded dataset. This data needs to be processed sequentially and incrementally on a event-by-event basis. Streaming data varies widely, from logs generated by customers using mobile or web applications, sensor data from Internet-of-Things (IoT) devices to credit card transactions. These examples of infinite datasets contrast with batch processing which works on finite ones. This distinction is important since some aggregations are very hard and even impossible to compute with 100\% accuracy for several use cases in an efficient manner on unbounded datasets. An example of such is counting distinct elements that flow through the data stream. An immediate solution would be to create a regular set of items or strings and have its length be the number of distinct elements processed. However, this approach is not very efficient and has a linear memory usage growth regarding the number of distinct elements. Therefore, as the number of distinct elements approaches large orders of magnitude we run out of memory.

@@

The Venn diagram \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques. A data stream is unbounded or infinite while a bounded data stream has a beginning and an end. Using these definitions, the concept of an unbounded data stream contains that of a bounded dataset. That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite dataset. This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.

@@

In Section 2.2 we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.

@@

In Section 2.1 we have defined a data stream as an unbounded dataset. But what exactly are the issues of working with an infinite stream of data? Why might we need to convert it into a bounded dataset and how to do so? In this Section, we will analyze such issues and present sliding windows as the solution.

@@

\subsection{Sliding Windows}

Botan et. al \cite{Botan-SECRET} formally define a window as:

\begin{definition}
A window W over a stream S is a finite subset of S"
\end{definition}

% 2 dimensions of windows
Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$. Windows can be defined as tuple or time-based. Tuple-based windows are windows whose size is defined by \emph{number} of items contained in the window, whereas time-based windows are windows whose size is defined by the total \emph{time} a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is formally defined as the distance between consecutive windows \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

@@

\subsection{Sliding Window Aggregations}

@@

\subsection{Sliding Window Aggregation Algorithms} \label{sec:back-swag-algs}

@@

Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011} \cite{EMA-Hunter} \cite{EMA-MarcusB} are a type of reduce functions applied to a sequence of elements, one at a time in a orderly fashion, to compute a weighted average, using different weights for each element. These weights decrease exponentially with the age of the observations. In other words, EMAs give higher weights to more recent data points. Figures \ref{fig:window-weights} and \ref{fig:ema-weights} show the weights applied to past events. In a typical sliding window, events either have weight of one (inside the window) or zero (outside the window) whereas EMAs give exponentially smaller weights to older events.

@@

Note that it is also possible to express a tuple-based half-life $n_{1/2}$, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half. In this case, the definition does not take into account the time passed between events:
\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $2^{-\frac{1}{n_{1/2}}}$ + val\textsubscript{t}
\label{def:tuple-ema}
\end{definition}
For example, with a tuple-based half-life of $\tau = 1$ tuple, for each new tuple all weights are reduced by half.

Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones. In this Thesis we chose to work with tuple-based EMAs due to its simplicity but our approach can be adjusted to use time-based ones.

>>>>

In this chapter we discuss fundamental concepts for our working context. In Section \ref{sec:stream-superset} we introduce batch and stream processing as two large scale data processing techniques. In Section \ref{sec:windows} we review windowing mechanisms that essentially turn unbounded data streams to finite datasets. In Section \ref{sec:aggregations} we introduce aggregations as a whole. Then, we specifically discuss sliding window aggregations, algorithms used for its computation and aggregation properties. In Section \ref{sec:emas} we present Exponential Moving Averages as aggregators that replace the sliding window contents for a lighter representation, providing an approximate estimation of the aggregation value.

@@

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch datasets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes). Between batch and stream processing, batch processing is the older paradigm. Historically, Apache Hadoop \cite{borthakur2007hadoop, Hadoop, ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark, Spark} was released as the industry accepted successor.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

Stream processing, on the other hand, refers to the processing of data streams. A data stream is essentially a continuous or unbounded dataset. This data needs to be processed sequentially and incrementally on a event-by-event basis. Streaming data varies widely, from logs generated by customers using mobile or web applications, sensor data from Internet-of-Things (IoT) devices to credit card transactions. These examples of infinite datasets contrast with batch processing which works on finite ones. This distinction is important since some aggregations are very hard and even impossible to compute with 100\% accuracy for several use cases in an efficient manner on unbounded datasets. An example of such is counting distinct elements that flow through the data stream. An immediate solution would be to create a regular set of items or strings and have its length be the number of distinct elements processed. However, this approach is not very efficient and has a linear memory usage growth regarding the number of distinct elements. Therefore, as the number of distinct elements approaches large orders of magnitude we run out of memory.

@@

The Venn diagram in Figure \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques. A data stream is unbounded or infinite while a bounded data stream has a beginning and an end. Using these definitions, the concept of an unbounded data stream contains that of a bounded dataset. That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite dataset. This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.

@@

In Section \ref{sec:windows} we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.

@@

In Section \ref{sec:stream-superset} we have defined a data stream as an unbounded dataset. But what exactly are the issues of working with an infinite stream of data? Why might we need to convert it into a bounded dataset and how to do so? In this section, we will analyze such issues and present sliding windows as the solution.

@@

Botan et. al \cite{Botan-SECRET} define a window as:
\begin{definition}
A window W over a stream $\mathbb{S}$ is a finite subset of $\mathbb{S}$
\end{definition}
Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$. Windows can be defined as tuple or time-based. Tuple-based windows are windows whose size is defined by \emph{number} of items contained in the window, whereas time-based windows are windows whose size is defined by the total \emph{time} a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is defined as the distance between consecutive windows \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

@@

\subsection{Sliding window aggregations}

@@

\subsection{Sliding window aggregation algorithms} \label{sec:back-swag-algs}

@@

Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB} are a type of reduce functions applied to a sequence of elements, one at a time in a orderly fashion, to compute a weighted average, using different weights for each element. These weights decrease exponentially with the age of the observations. In other words, EMAs give higher weights to more recent data points. Figures \ref{fig:window-weights} and \ref{fig:ema-weights} show the weights applied to past events. In a typical sliding window, events either have weight of one (inside the window) or zero (outside the window) whereas EMAs give exponentially smaller weights to older events.

@@

Note that it is also possible to express a tuple-based half-life $n_{1/2}$, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half. In this case, the definition does not take into account the time passed between events:
\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $2^{-\frac{1}{n_{1/2}}}$ + val\textsubscript{t}
\label{def:tuple-ema}
\end{definition}
For example, with a tuple-based half-life of $n_{1/2}=1$ tuple, for each new tuple all weights are reduced by half.

Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones. In this Thesis we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.

### 2020-07-02 chapters/introduction.tex
Restivo feedback applied

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized in terms of its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former is best known as stream processing, whereas the latter is known as batch processing.

@@

In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations that produce a single result when applied to data sets are commonly known as aggregation operations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis of stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set typically by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.

Consider the use-case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day. In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one. Such a system would have to process all incoming transactions and decide in a fraction of a second, for each transaction, whether it is fraudulent or not. It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.

Many market solutions monitor streams of credit card transactions with the intent of fraud detection. In the context of this Thesis, we work side by side with Feedzai\footnote{https://feedzai.com}. In brief, the company's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained,  before deployment, often with assumptions on the statistical distributions of future incoming data. Thus their model performance depends on how well those assumptions hold for future data. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.} Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.} new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift \cite{drift} \cite{gama-drift}.

@@

\section{Hypothesis} \label{sec:hypothesis}
The question this Thesis aims to answer is whether sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint. In this Thesis, we study some sliding window aggregations that do not keep the sliding window in memory but still give an estimate on the aggregation value.

@@

The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item low memory footprint

@@

    \item explainable alerts
\end{itemize}

Experiments will be made using multiple synthetic data sets and real ones from the financial fraud space.

>>>>

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized by its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former is best known as stream processing, whereas the latter is known as batch processing.

@@

In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations that produce a single result when applied to data sets are commonly known as aggregation operations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size --- \textit{e.g.}, events from the last three days or the last million events, respectively.

Consider the use-case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day. In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one. Such a system would have to process all incoming transactions and decide, in a fraction of a second, if a transaction is fraudulent or not. It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.

Many market solutions monitor streams of credit card transactions with the intent of fraud detection. In the context of this Thesis, we work side by side with Feedzai\footnote{https://feedzai.com}. In brief, the company's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained,  before deployment, often with assumptions on the statistical distributions of future incoming data. Thus their model performance depends on how well those assumptions hold for future data. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.}, Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.}, new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift \cite{drift, gama-drift}.

@@

\section{Hypothesis} \label{sec:hypothesis}
The question this Thesis aims to answer is whether \textit{sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.} In this Thesis, we study sliding window aggregations that replace the sliding window events in memory by a lighter representation, while still being able to provide an estimate of the aggregation value.

@@

The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item low memory footprint

@@

    \item explainable alerts
\end{itemize}
The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic data sets and a real dataset from the financial fraud space.

### 2020-07-02 chapters/problem-statement.tex
Restivo feedback applied

With this Chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.

\section{The problem}
Many real-time stream monitoring systems are static once deployed in a production environment. Engineers of said systems configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data. They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system. Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates. The problem at hand is to determine when to reconfigure the system, \textit{e.g.} a Machine Learning model, based on streaming analysis.

@@

    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout run-time

@@

\section{Proposal}
We devise a two-phased (batch and streaming phases) and two-windowed (reference and target sliding windows) method. We devise a sliding window approximate histogram aggregation that encodes data distributions and provide a method for incrementally maintaining it with constant time and space complexity for every growing volumes of data, ideal for the streaming scenario. Given a user-defined reference period and a stream of data to process, we monitor each event's fields and alert if they are considered divergent relative to their reference distribution. We provide a measure of how divergent each field (or feature) is. Based on this information, it is our belief that a system administrator would have an easier time determining when to reconfigure the system.

@@

\section{Research Questions}
The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time. To that end, we devise a series of research questions to be answered by the end of this Thesis:

>>>>

With this chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.

\section{The problem}
Many real-time stream monitoring systems are static once deployed in a production environment. The engineers of those systems configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data. They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system. Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates. The problem at hand is to determine when to reconfigure the system, \textit{e.g.} a Machine Learning model, based on an analysis of the drifts in the stream of data.

@@

    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout runtime

@@

\section{Proposal}
%2 phases, names and purpose
We devise a two-phased and two-windowed method: the first phase is a batch analysis that makes use of a reference window; and the second phase is a streaming analysis that makes use of a target sliding window. The batch analysis main goal is to compute the reference aggregation for the reference window. The stream analysis main goals are to incrementally maintain the same type of aggregation but over the streaming data and alert when the difference between the reference aggregation and the target sliding one is high.

%aggregations to represent distributions
We choose to use an histogram aggregation with constant memory usage to encode the distribution of the observed values. Furthermore, we devise a sliding window approximate histogram aggregation that is maintainable in constant time in a sliding window fashion, through the use of Exponential Moving Averages (EMAs). Hence, our EMA-like histogram aggregation is ideal for the streaming scenario because of its constant time and space complexity, for every growing volumes of data.

Given a user-defined reference period and a stream of data to process, we monitor each event's fields and alert if they are considered divergent relative to their reference aggregation. We provide a measure of how divergent each field is. Based on this information, it is our belief that a system administrator would have an easier time determining when to reconfigure the system.

@@

\section{Research questions}
The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time. To that end, we devise a series of research questions to be answered by the end of this Thesis:

\textcolor{red}{TODO: further describe each RQ?}
